\documentclass[journal]{IEEEtran}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{orcidlink}

\begin{document}

\title{Prequential Streaming Evaluation of Stateful Time-Series Forecasters with
Online Conformal Calibration}

\author{%
  Felipe Santib\'a\~nez-Leal~\orcidlink{0000-0002-0150-3246},~\IEEEmembership{Independent Researcher}%
  \thanks{F.\ Santib\'a\~nez-Leal is with the CAOS open-research programme,
    Santiago, Chile (e-mail: \texttt{fsantibanezleal@ug.uchile.cl};
    ORCID: 0000-0002-0150-3246).}%
  \thanks{Preprint, 2026. The harness is released as the open-source package
    \texttt{preqts} (\url{https://pypi.org/project/preqts/}); the atlas testbed
    (code, committed per-case artifacts, interactive workbench) is at
    \url{https://github.com/fsantibanezleal/CAOS_RES_ChronoScope} (MIT), live at
    \url{https://chronoscope.fasl-work.com}.}%
}

\markboth{Preprint, 2026 --- CAOS open-research programme}{Santib\'a\~nez-Leal:
Prequential Streaming Evaluation of Stateful Forecasters with Online Conformal Calibration}

\IEEEtitleabstractindextext{%
\begin{abstract}
Real forecasting is streaming: observations arrive one by one, the model updates,
and its prediction intervals must stay calibrated as the series drifts. Yet every
widely used forecasting benchmark harness evaluates models statelessly on rolling
windows, carrying no state across windows and never running a test-then-train loop.
We address this verified gap with a prequential streaming evaluation harness for
\emph{stateful} forecasters, released as the open-source package \texttt{preqts}: it
implements Dawid's predict-then-observe-then-update protocol behind a
\texttt{StatefulForecaster} protocol with an explicit covariate-arrival policy, and
records four trajectories per method, rolling scaled error, rolling empirical
coverage against the nominal level, interval width, and cumulative compute cost, the
last of which separates a genuinely stateful constant-cost model from a
replay-everything wrapper. We use the harness to test a concrete claim: that online
conformal calibration keeps streaming intervals at their nominal level under
distribution shift. On an atlas of fifteen series spanning distinct regimes (level
shift, chaos, heteroskedasticity, intermittency, long memory, multi-seasonality,
exogenous promotions, and real UCI/M4/PM2.5 data) plus white-noise and random-walk
honesty controls, wrapping a Theta point forecaster in Adaptive Conformal Inference
or Conformal-PID control reduces the mean absolute coverage error at the nominal
$80\%$ level from $0.042$ (raw Theta) and $0.060$ (seasonal-naive) to $0.0057$ and
$0.0060$ respectively, a roughly seven-to-ten-fold reduction, and caps the
worst-case per-series coverage error at $0.022$ versus $0.11$--$0.31$ uncalibrated,
without changing the point forecast. The calibration holds on the honesty controls,
where a spurious skill gap would flag harness leakage. The contribution is the
reusable stateful-streaming harness and a leakage-guarded, artifact-backed
demonstration of where online conformal calibration earns its keep.
\end{abstract}

\begin{IEEEkeywords}
time-series forecasting, prequential evaluation, streaming, conformal prediction,
adaptive conformal inference, calibration, coverage, stateful forecasters,
foundation models, benchmark, reproducibility.
\end{IEEEkeywords}}

\maketitle
\IEEEdisplaynotcompsoctitleabstractindextext
\IEEEpeerreviewmaketitle

% =====================================================================
\section{Introduction}
% =====================================================================

\IEEEPARstart{F}{orecasting} in production is a streaming problem: a datum arrives,
the model issues a prediction for the next horizon, the truth is later revealed, and
the model updates its state before the next step. The prediction interval must remain
calibrated throughout, even as the series drifts across regimes. This is not how
forecasting models are usually evaluated. The widely used benchmark harnesses,
including the rolling-window backtests of GluonTS and Darts and the
window-based fev-bench~\cite{shchur2025}, score forecasters \emph{statelessly}: each
window is independent, no learned state carries across windows, and there is no
test-then-train loop. The verified gap is that no public library evaluates a
\emph{stateful} streaming forecaster prequentially with an explicit policy for when
covariates arrive.

We fill it with \texttt{preqts}, an open-source harness that implements the
prequential principle, Dawid's ``predictive sequential'' test-then-train
protocol~\cite{dawid1984}: at each step the model predicts, the truth is revealed and
scored, and only then does the model update. Skill becomes a trajectory over the
stream rather than a single number, an adapting model shows a falling error curve
while a static one drifts. The harness sits behind a \texttt{StatefulForecaster}
protocol so that classical methods, gradient-boosted models, and stateful foundation
models are all evaluated identically, and it records the compute cost per step so
that a constant-cost stateful model is visibly distinguished from a wrapper that
re-reads the whole history each step.

We then use the harness to answer a specific, practical question: does online
conformal calibration keep streaming intervals calibrated under distribution shift?
Conformal prediction offers distribution-free coverage guarantees, and two online
variants, Adaptive Conformal Inference (ACI)~\cite{gibbs2021} and Conformal-PID
control~\cite{angelopoulos2023}, are designed for exactly the drifting-stream
setting. We measure their effect on an atlas of fifteen series that deliberately
spans distinct forecasting regimes and includes white-noise and random-walk honesty
controls, and we report the result artifact-backed and leakage-guarded. Our
contributions are: (i) the reusable prequential streaming harness \texttt{preqts}
for stateful forecasters with a covariate-arrival policy; (ii) an artifact-backed,
control-guarded demonstration that online conformal calibration reduces streaming
coverage error roughly seven-to-tenfold and caps the worst case, without changing
the point forecast.

% =====================================================================
\section{The prequential streaming harness}
\label{sec:harness}
% =====================================================================

\subsection{Protocol}

Let a stream deliver observations $y_1, y_2, \dots$ At step $t$ a stateful forecaster
$f$ with state $s_t$ issues a horizon-$h$ prediction and an interval at nominal level
$1-\alpha$; the truth $y_{t+h}$ is then revealed, the prediction is scored, and the
state updates:
\begin{equation}
(\hat{y}_{t+h}, \hat{I}_t) = f_{\text{predict}}(s_t), \qquad
s_{t+1} = f_{\text{update}}(s_t, y_{t+h}).
\end{equation}
The update strictly follows scoring, so no point is ever used to predict itself; a
covariate-arrival policy declares, per exogenous channel, whether its value at $t+h$
is known at prediction time (a planned promotion) or only revealed with the target
(a realized driver). After a warmup of $w$ steps the harness records, as functions of
stream position, four trajectories: the rolling scaled error, the rolling empirical
coverage, the interval width, and the cumulative compute cost.

\subsection{Metrics}

Point accuracy uses the mean absolute scaled error (MASE)~\cite{hyndman2006}, which
divides the absolute error by the in-sample seasonal-naive error so that values below
one beat the naive baseline and the scale is comparable across series. Interval
quality is read two ways that cannot be gamed independently: the empirical coverage
(the fraction of revealed truths inside the interval, compared to the nominal
$1-\alpha$) and the interval width (a calibrator that only widens to reach coverage is
not impressive), and both are summarized by the M4 interval score MSIS~\cite{makridakis2020},
which charges width plus a miss penalty per unit so a narrow interval that misses
cannot look good. Probabilistic sharpness is scored with the weighted quantile loss,
a proper scoring rule~\cite{gneiting2007}. The compute-cost trajectory is the
constant-cost-per-step signature that separates a stateful model from a
replay-everything wrapper.

\subsection{The \texttt{preqts} package}

The harness is extracted from the atlas and published as the open-source package
\texttt{preqts}. It provides the \texttt{StatefulForecaster} protocol, the streaming
loop with the covariate-arrival policy, the four-trajectory recorder, and the online
conformal calibrators of Sec.~\ref{sec:conformal}. Because the streaming artifact
carries only aggregate metric trajectories (no raw series values are recoverable from
rolling MASE, coverage, width, or cost), the demonstration ships even for
local-only-licensed real datasets whose raw excerpts are otherwise redacted.

% =====================================================================
\section{Online conformal calibration}
\label{sec:conformal}
% =====================================================================

Given a nonconformity score (here the absolute residual), split-conformal prediction
issues an interval whose width is a quantile of past scores; under exchangeability its
coverage is $1-\alpha$. Streams are not exchangeable, so the effective level must
adapt. \textbf{Adaptive Conformal Inference}~\cite{gibbs2021} updates the level each
step by the observed miss,
\begin{equation}
\alpha_{t+1} = \alpha_t + \gamma\,\big(\alpha - \mathrm{err}_t\big),
\qquad \mathrm{err}_t = \mathbf{1}\{y_{t+h}\notin \hat{I}_t\},
\label{eq:aci}
\end{equation}
so that long-run coverage converges to the target under arbitrary distribution shift.
\textbf{Conformal-PID control}~\cite{angelopoulos2023} generalizes this to a
proportional-integral-derivative controller on the coverage error; ACI is the
integral-only special case, and the proportional and derivative terms buy faster
coverage recovery after a regime change. Both wrap any point forecaster and change
only the interval, never the point prediction.

% =====================================================================
\section{The atlas testbed}
\label{sec:atlas}
% =====================================================================

We exercise the harness on ChronoScope, an atlas of fifteen committed cases baked
deterministically (each artifact a pure function of case and seed). The cases span
distinct regimes by construction, level shift, deterministic chaos (Mackey--Glass),
GARCH heteroskedasticity, intermittent demand, long memory (fractional), coupled
daily/weekly seasonality, and exogenous promotions, and include real CC-BY series
(UCI Electricity, Beijing PM2.5, M4 daily and hourly) and two honesty controls,
white noise and a near-random walk, on which a suspiciously large ``skill'' would flag
harness leakage. Each case is scored by the full method ladder (classical, statistical,
gradient-boosted, deep, and stateful foundation models such as
Chronos~\cite{ansari2024} and TimesFM~\cite{das2024}) and by the streaming bench of
Sec.~\ref{sec:harness} at nominal coverage $80\%$, horizon one, warmup twenty-four.
The streaming roster reported here is the honest floor (seasonal-naive), a strong
classical live method (Theta~\cite{assimakopoulos2000}) with raw intervals, and that
same Theta wrapped in ACI and in Conformal-PID.

% =====================================================================
\section{Results}
% =====================================================================

Table~\ref{tab:coverage} is the headline. Averaged over the fifteen cases, the raw
Theta intervals miss the nominal $80\%$ target by a mean absolute coverage error of
$0.042$ and the seasonal-naive floor by $0.060$, with worst-case per-series errors of
$0.11$ and $0.31$. Wrapping the \emph{same} Theta point forecaster in ACI or
Conformal-PID cuts the mean coverage error to $0.0057$ and $0.0060$, a seven-fold
reduction over raw Theta and a ten-fold reduction over seasonal-naive, and caps the
worst-case per-series error at $0.022$. The point forecast is unchanged (the MASE of
Theta, Theta+ACI and Theta+PID is identical per case); only the interval self-corrects.

\begin{table}[!t]
\caption{Streaming interval calibration at nominal $80\%$ over 15 cases (mean and
worst-case absolute coverage error $|\,\text{coverage}-0.80\,|$). Online conformal
calibration cuts the error $7$--$10\times$ and caps the worst case.}
\label{tab:coverage}
\centering
\begin{tabular}{lrr}
\toprule
Streaming method & mean $|\text{cov}-0.80|$ & worst case \\
\midrule
Seasonal-naive (floor) & 0.0599 & 0.307 \\
Theta, raw intervals & 0.0416 & 0.112 \\
Theta + ACI & \textbf{0.0057} & \textbf{0.022} \\
Theta + Conformal-PID & \textbf{0.0060} & \textbf{0.022} \\
\bottomrule
\end{tabular}
\end{table}

The effect is regime-robust. Where raw intervals drift most, the calibrators pull
coverage back to target: on intermittent demand the raw Theta over-covers at $0.91$
while ACI sits at $0.807$; on a chaotic Mackey--Glass stream raw Theta under-covers at
$0.77$ while ACI reaches $0.801$; on real UCI electricity raw Theta under-covers at
$0.76$ while ACI reaches $0.796$. On the white-noise and random-walk honesty controls
the calibrated coverage holds at $0.79$ (against raw $0.75$), i.e.\ the harness does
not manufacture skill on unpredictable series, exactly the leakage guard the controls
exist to enforce. Because the point forecast never changes, this is a pure calibration
gain: the same forecaster, with intervals that stay honest as the stream drifts.

% =====================================================================
\section{Discussion and honest scope}
% =====================================================================

The finding, that online conformal calibration keeps streaming intervals near nominal
under shift, is not by itself surprising: it is what ACI and Conformal-PID were
designed to do. The contribution is that it is measured under the correct protocol.
A stateless rolling-window harness cannot see a calibrator recover coverage after a
regime change, because it does not carry the calibrator's state across windows; the
prequential harness does, and it charges the compute cost that distinguishes a
constant-cost stateful method from a replay wrapper. The honesty controls and the
identical point forecasts across the raw and calibrated variants rule out the two
easy ways such a result could be an artifact (harness leakage; a point-accuracy
confound).

Scope. The demonstration uses a Theta base forecaster because it is a strong classical
method that runs live; the calibration is method-agnostic and wraps any forecaster,
which is the point of the protocol. The atlas mixes seeded synthetic regimes with a
few real CC-BY series; the synthetic cases are constructed to isolate regimes and the
controls guard against leakage, but the real-data coverage is reported on a handful of
public series, not a large corpus. Results are shown at horizon one and nominal $80\%$;
the harness supports other horizons and levels. The value we claim is the reusable
stateful-streaming harness and the leakage-guarded measurement, not a new calibrator.

% =====================================================================
\section{Conclusion}
% =====================================================================

We built and released \texttt{preqts}, a prequential streaming evaluation harness for
stateful time-series forecasters with an explicit covariate-arrival policy, filling a
verified gap left by stateless-window benchmarks, and used it to show, artifact-backed
and control-guarded, that online conformal calibration reduces streaming interval
coverage error roughly seven-to-tenfold and caps the worst case while leaving the
point forecast unchanged. Evaluating forecasters the way they are deployed, as a
stream with state, is what makes the calibration story visible; the harness and the
committed atlas make it reproducible.

% =====================================================================
\section*{Data and code availability}
% =====================================================================

The harness is the open-source package \texttt{preqts} (\url{https://pypi.org/project/preqts/}).
The atlas code, the committed per-case artifacts (trace, analysis, and streaming
trajectories) and the interactive workbench are at
\url{https://github.com/fsantibanezleal/CAOS_RES_ChronoScope} under the MIT license;
live at \url{https://chronoscope.fasl-work.com}. Real datasets (UCI Electricity,
Beijing PM2.5, M4 excerpts) ship under CC-BY-4.0 with attribution; streaming artifacts
carry aggregate trajectories only.

% =====================================================================
\begin{thebibliography}{99}
\itemsep 0pt

\bibitem{dawid1984} A.~P. Dawid, ``Present position and potential developments: Some
personal views. Statistical theory. The prequential approach,'' \emph{J. R. Stat.
Soc. A}, vol.~147, no.~2, pp.~278--292, 1984, doi:10.2307/2981683.

\bibitem{gibbs2021} I.~Gibbs and E.~Cand\`es, ``Adaptive conformal inference under
distribution shift,'' in \emph{Adv. Neural Inf. Process. Syst. (NeurIPS)}, 2021,
arXiv:2106.00170.

\bibitem{angelopoulos2023} A.~N. Angelopoulos, E.~J. Cand\`es, and R.~J. Tibshirani,
``Conformal PID control for time series prediction,'' in \emph{Adv. Neural Inf.
Process. Syst. (NeurIPS)}, 2023, arXiv:2307.16895.

\bibitem{shchur2025} O.~Shchur \emph{et al.}, ``fev-bench: A realistic benchmark for
time series forecasting,'' 2025, arXiv:2509.26468.

\bibitem{hyndman2006} R.~J. Hyndman and A.~B. Koehler, ``Another look at measures of
forecast accuracy,'' \emph{Int. J. Forecast.}, vol.~22, no.~4, pp.~679--688, 2006,
doi:10.1016/j.ijforecast.2006.03.001.

\bibitem{makridakis2020} S.~Makridakis, E.~Spiliotis, and V.~Assimakopoulos, ``The M4
competition: 100,000 time series and 61 forecasting methods,'' \emph{Int. J.
Forecast.}, vol.~36, no.~1, pp.~54--74, 2020, doi:10.1016/j.ijforecast.2019.04.014.

\bibitem{gneiting2007} T.~Gneiting and A.~E. Raftery, ``Strictly proper scoring rules,
prediction, and estimation,'' \emph{J. Amer. Statist. Assoc.}, vol.~102, no.~477,
pp.~359--378, 2007, doi:10.1198/016214506000001437.

\bibitem{assimakopoulos2000} V.~Assimakopoulos and K.~Nikolopoulos, ``The theta model:
A decomposition approach to forecasting,'' \emph{Int. J. Forecast.}, vol.~16, no.~4,
pp.~521--530, 2000, doi:10.1016/S0169-2070(00)00066-2.

\bibitem{ansari2024} A.~F. Ansari \emph{et al.}, ``Chronos: Learning the language of
time series,'' \emph{Trans. Mach. Learn. Res.}, 2024, arXiv:2403.07815.

\bibitem{das2024} A.~Das, W.~Kong, R.~Sen, and Y.~Zhou, ``A decoder-only foundation
model for time-series forecasting,'' in \emph{Proc. Int. Conf. Mach. Learn. (ICML)},
2024, arXiv:2310.10688.

\end{thebibliography}

\end{document}
